---
title: "Homework 4"
author: "Xinyao Yin"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: yes
    toc: true
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{The ridge regression vignette}
-->

##Question 1: CASL 5.8.2
According to CASL Equation 5.19, the Hessian matrix can be written as the inner-product of the matrix X weighted by the variance of each observation
$$H(\ell) = X^t\textrm{diag}(p(1-p))X$$

According to CASL Equation 2.73, 
$$cond(A)=\frac{\sigma_{max}}{\sigma_{min}}$$

Generate matrix X 
```{r}
X <- matrix(c(3, 1, 1, 3), 2, 2)
```

$$
X = \begin{pmatrix}
3 & 1 \\
1 & 3 \\
\end{pmatrix}
$$
For probability p, we know that the function f(p)=p(1-p) has a maximum value of 0.25 at p=0.5 and gradually goes to 0 as $p\to0$ and $p\to1$. So if we set $p_1=0.5$ and $p_2=10^{-8}$ the logistic variation of the Hessian Matrix will be poorly-conditioned.

Generate probability p 
```{r}
p <- c(0.5, 0.00000001)
```

$$
p = (0.5, 0.00000001)
$$

The condition number of the linear Hessian ($X'X$) is:
```{r}
svals <- svd(t(X)%*%X)[["d"]]
max(svals)/min(svals)
```

The logistic variation of the Hessian ($X'X$) is: 
$$X^t\textrm{diag}(p(1-p))X$$

The condition number of the logistic variation of the Hessian is: 
```{r}
H = t(X)%*%diag(p*(1-p))%*%X
svals2 <- svd(H)[["d"]]
max(svals2)/min(svals2)
```

As we can see that the linear Hessian ($X^tX$) is well-conditioned with a condition number of 4, but the logistic variation Hessian matrix $X^t\textrm{diag}(p(1-p))X$ is ill-conditioned with a much larger conditional number of 39062501. 


##Question 2: CASL 5.8.4

```{r}
# Based on modification of the irwls_glm function on CASL Page 127

# Inputs: X: a numeric design matrix; 
#         y: a numeric vector;
#         family: an instance of a family object;
#         maxit: integer maximum iterations; 
#         tol: stopping criterion tolerance
#         lambda: tunning parameter of the l-2 norm penalty 
# Output: estimate of the GLM regression vector

irwls_glm <- function(X, y, family, maxit=25, tol=1e-10, lambda){
  beta <- rep(0,ncol(X))
  for(j in 1:maxit){
    b_old <- beta
    eta   <- X %*% beta
    mu    <- family$linkinv(eta)
    mu_p  <- family$mu.eta(eta)
    z     <- eta + (y - mu) / mu_p
    W     <- as.numeric(mu_p^2 / family$variance(mu))
    XtX   <- crossprod(X, diag(W) %*% X)
    Xtz   <- crossprod(X, W * z)
    
    # add l-2 norm penalty 
    L <- XtX + diag(lambda, dim(XtX)[1])
    beta  <- solve(L, Xtz)
    if(sqrt(crossprod(beta - b_old)) < tol) break
  }
  beta
}
```

##Question 3: see sparse_matrix.r in the R folder. 